---
title: "Data Dictionary"
author: "Tracy Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(janitor)
library(dplyr)
library(tidyverse)
```

Read CSV File
```{r}
data <- read.csv("lc_loan.csv")
head(data)
```
```{r}
data <- clean_names(data)
```

```{r}
names(data)
```


Subset of the actual data
Numerical variables: annual_income, int_rate, loan_amnt, emp_length, dti
Categorical variables: addr_state, grade, purpose, home_ownership, term

```{r}
actual_data <- subset(data, select = c("annual_inc", "int_rate", "loan_amnt", "emp_length", "dti", "addr_state", "grade", "purpose", "home_ownership", "term"))
```

```{r}
write.csv(actual_data, "subset_data.csv", row.names = FALSE)
```

```{r}
glimpse(actual_data)
```

Data Dictionary:

| Variable | Type | What it represents|
| :--- | :---: | :---: |
| int_rate | numeric (double) | interest rate on the loan |
| loan_amnt | numeric (double) | amount of the loan issued |
| annual_inc | numeric (double) | annual income of the borrower |
| emp_length | categorical | employment length in years |
| dti | numeric (double) | debt-to-income ratio |
| grade | categorical | loan grade assigned by Lending Club (A-G|
| purpose | categorical | stated purposed of the loan |
| home_ownership | categorical | Borrower's home ownership status |
| addr_state | categorical | state of the borrower |
| term | categorical | length of the loan (36 or 60 months) |


Questions:

1. Can interest rate be predicted based on loan term and loan grade?
boxplot of int_rate by grade, faceted by term (36 vs 60 months)

```{r}

```


2. Can home ownership status be predicted using borrower income and employment length?
boxplot of annual_income by home_ownership, plus a grouped bar of emp_length x home_ownership

```{r}

```


3. How do average interest rates for each grade compare?
- This is one numeric and one categorical variable.
Summary table: grouped means of int_rate by grade
Graph: boxplot of int_rate by grade (shows distribution + outliers per grade, not just averages)

```{r}

```


4. Does purpose of loan correlate to loan amount?
- One numeric and one categorical
Summary table: grouped stats of loan_amnt by purpose
Graph: boxplot

```{r}

```


5. Which states have the lowest average grades?
- One numeric/ordinal + one categorical
Summary table: average grade by state, sorted from lowest to highest
Graph: bar chart of average grade by state, sorted ascending

```{r}

```


6. How does employment length relate to home ownership status?
- Two categorical variables
Summary table: table(df$emp_length, df$home_ownership)
Graph: grouped bar chart of home ownership proportions across employment length groups

```{r}

```


7. Is there a relationship between state and purpose of loan
- Two categorical variables
Summary table: contingency table of state x purpose
Graph: stacked bar chart of purpose by state (each bar is a state, segments are purposes

```{r}

```


8. Is there a relationship between loan amount and loan grade?
- One numeric and one categorical
Summary table: grouped stats of loan_amnt by grade using profiling_num() after grouping
Graph: boxplot of loan_amnt by grade

```{r}

```


9. Is there a correlation between DTI and purpose?
Summary table: grouped means/medians of DTI by purpose
Graph: boxplot of DIT by purpose

```{r}

```


10. Is there a relationship between home ownership and purpose?
- Two categorical variables
Summary table: table(df$home_ownership, df$purpose)
Graph: grouped/stacked bar chart showing proportions of loan purpose within each home ownership category

```{r}

```


Additional changes to replace:

