---
title: "Exploratory Data Analysis"
author: "STOR 320.(01 OR 02) Group PLACE_GROUP_NUMBER_HERE (Ex: STOR 320.01 Group 12)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

### Statement of business problem



### Data Dictionary:
| Variable | Type | What it represents|
| :--- | :---: | :---: |
| int_rate | numeric (double) | interest rate on the loan |
| loan_amnt | numeric (double) | amount of the loan issued |
| annual_inc | numeric (double) | annual income of the borrower |
| emp_length | categorical | employment length in years |
| dti | numeric (double) | debt-to-income ratio |
| grade | categorical | loan grade assigned by Lending Club (A-G|
| purpose | categorical | stated purposed of the loan |
| home_ownership | categorical | Borrower's home ownership status |
| addr_state | categorical | state of the borrower |
| term | categorical | length of the loan (36 or 60 months) |


### Data Quality Report:
Numeric data quality report:



Categorical data quality report:


### Data Quality Plan:




### Data Cleaning
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# Import Data Below
loan <- read_csv("../Raw Data/lc_loan.csv")
```

```{r}
library(janitor)      # for clean_names()
library(stringr)      # for str_squish()
library(DataExplorer) # for introduce() and plot_missing()
```

```{r}
glimpse(loan)
```

```{r}
dim(loan)
```

```{r}
DataExplorer::introduce(loan)
DataExplorer::plot_missing(loan)
```

```{r}
(missing_tbl <- loan |>
    summarise(across(everything(), ~ sum(is.na(.)))) |>
    pivot_longer(everything(), names_to = "Feature", values_to = "n_missing") |>
    arrange(desc(n_missing)))
any(duplicated(loan))
```

```{r}
loan <- loan[!duplicated(loan), ]

loan_clean <- loan |>
  select(
    annual_inc, int_rate, loan_amnt, emp_length, dti,   # numerical
    addr_state, grade, purpose, home_ownership, term,   # categorical
    loan_status, verification_status                    # useful context
  )

loan_clean <- loan_clean |>
  clean_names()

loan_clean <- loan_clean |>
  mutate(across(where(is.character), ~ str_squish(.))) |>
  mutate(across(where(is.character), ~ na_if(.x, ""))) |>
  mutate(across(where(is.character),
                ~ case_when(
                  .x %in% c("NA", "N/A", "null", "NULL", "?", "unknown", "Unknown") ~ NA_character_,
                  TRUE ~ .x
                )))

loan_clean <- loan_clean |>
  mutate(across(all_of(c("grade", "addr_state", "home_ownership")), str_to_upper)) |>
  mutate(across(all_of(c("purpose", "term", "loan_status", "verification_status")), str_to_lower))
```

```{r}
loan_clean <- loan_clean |>
  mutate(emp_length = suppressWarnings(case_when(
    emp_length == "< 1 year"  ~ 0,
    emp_length == "1 year"    ~ 1,
    emp_length == "10+ years" ~ 10,
    TRUE ~ as.numeric(gsub(" years?", "", emp_length))
  )))

loan_clean <- loan_clean |>
  mutate(
    annual_inc = as.numeric(annual_inc),
    loan_amnt  = as.numeric(loan_amnt),
    int_rate   = as.numeric(gsub("%", "", int_rate)),
    dti        = as.numeric(dti)
  )

loan_clean <- loan_clean |>
  mutate(
    grade          = factor(grade, levels = c("A","B","C","D","E","F","G"), ordered = TRUE),
    term           = factor(term,  levels = c("36 months","60 months"),     ordered = TRUE),
    home_ownership = factor(home_ownership,
                            levels = c("OWN","MORTGAGE","RENT","OTHER","NONE","ANY")),
    purpose        = factor(purpose),
    addr_state     = factor(addr_state)
  )
```

```{r}
table(factor(loan_clean$grade))
```

```{r}
table(factor(loan_clean$home_ownership))
```

```{r}
table(factor(loan_clean$purpose))
```

```{r}
table(factor(loan_clean$term))
```

```{r}
loan_clean <- loan_clean |>
  filter(
    !is.na(int_rate),
    !is.na(loan_amnt),
    !is.na(grade),
    !is.na(home_ownership),
    !is.na(purpose),
    !is.na(term),
    !is.na(addr_state)
  )

glimpse(loan_clean)
dim(loan_clean)
```

```{r}
(missing_tbl_final <- loan_clean |>
    summarise(across(everything(), ~ sum(is.na(.)))) |>
    pivot_longer(everything(), names_to = "Feature", values_to = "n_missing") |>
    arrange(desc(n_missing)))
```

```{r}
loan_clean <- loan_clean |>
  mutate(purpose = str_replace_all(purpose, "_", " "))
```

```{r}
actual_data <- loan_clean |>
  select(annual_inc, int_rate, loan_amnt, emp_length, dti,
         addr_state, grade, purpose, home_ownership, term)
```







# Creator: Tracy Huang

### Q1: Can interest rate be predicted based on loan term and loan grade?

```{r}

```

### Q2: Can home ownership status be predicted using borrower income and employment length?

```{r}
#
```

# Interpreter: Kyran Lakhia

### Q1: Which states are experiencing the lowest grades on average?

```{r}
#
```

### Q2: Is there a relationship between home ownership and purpose?

```{r}
#
```

# Orator 1: Ross Guinan

### Q1: Is there a relationship between loan amount and loan grade?

```{r}
#
```

### Q2: Is there a correlation between DTI and purpose?

```{r}
#
```

# Orator 2: Carter Boxell

### Q1: How do the average interest rates for each grade compare?

```{r}
#
```

### Q2: Is there a relationship between state and purpose of loan?

```{r}
#
```

# Deliverer: Emily Jones

### Q1: Does the purpose of the loan correlate to the amount of the loan?

```{r}
#
```

### Q2: How does employment length relate to homeownership status?

```{r}
#
```

# Follow-up Questions

### New Questions Based Off Initial Investigation

- Q1: WRITE_QUESTION_HERE
- Q2: WRITE_QUESTION_HERE
- Q3: WRITE_QUESTION_HERE 
- Q4: WRITE_QUESTION_HERE

### Investigation of Follow-up Questions

GIVE WHAT 2 QUESTIONS YOU ATTEMPTED TO INVESTIGATE FURTHER IN COMPLETE SENTENCES (*Example:* Our group decided to investigate Q2 and Q4 in further detail.)

SHOW AT LEAST 2 TABLES OR FIGURES BELOW THAT EXPLORE ANSWERS FOR THE QUESTIONS YOU ARE INVESTIGATING FURTHER.

```{r}
#
```

```{r}
#
```

# Summary

GIVE A 2 PARAGRAPH SUMMARY. 

PARAGRAPH 1 SHOULD DESCRIBE WHAT YOU LEARNED ABOUT YOUR DATA FROM INVESTIGATING THE INITIAL QUESTIONS. DID YOU FIND ANYTHING UNUSUAL IN YOUR DATA? DID ANYTHING SURPRISE YOU? WHICH OF THE INITIAL QUESTIONS WERE HELPFUL IN LEADING YOU TO MORE QUESTIONS?

PARAGRAPH 2 SHOULD SUMMARIZE WHAT YOU LEARNED FROM INVESTIGATING THE FOLLOW-UP QUESTIONS. WHY ARE THESE FOLLOW-UP QUESTIONS INTERESTING FOR INVESTIGATION? DESCRIBE THE TABLES/FIGURES YOU USED TO EXPLORE ANSWERS TO THESE FOLLOW-UP QUESTIONS? WHAT DID YOU LEARN FROM THE TABLES/FIGURES REGARDING THE FOLLOW-UP QUESTIONS YOU PROPOSED?















