---
title: "Exploratory Data Analysis"
author: "STOR 320.(01 OR 02) Group PLACE_GROUP_NUMBER_HERE (Ex: STOR 320.01 Group 12)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressWarnings(library(tidyverse))
suppressWarnings(library(janitor))      
suppressWarnings(library(stringr))     
suppressWarnings(library(DataExplorer))
suppressWarnings(library(skimr))
suppressWarnings(library(corrr))
suppressWarnings(library(tidyverse))
suppressWarnings(library(car))
suppressWarnings(library(gt))
library(tidyr)
library(funModeling)
library(readr)
library(dplyr)
library(ggplot2)
```


```{r setup, include=FALSE}
# Import Data Below
loan_dataset_full <- read_csv("../Raw Data/lc_loan.csv")
```

Subset of actual dataset of interest:
```{r}
actual_data <- loan_dataset_full |>
  select(annual_inc, int_rate, loan_amnt, emp_length, dti,
         addr_state, grade, purpose, home_ownership, term) |>
  janitor::clean_names() |>
  mutate(across(where(is.character), as.factor))

glimpse(actual_data)
summary(actual_data)
```

###Statement of the business problem


### Data Dictionary
RESEARCH MORE*

| Variable | Type | What it represents |
| :--- | :---: | :--- |
| int_rate | Numeric (continuous) | Interest rate on the loan |
| loan_amnt | Numeric (continuous) | Amount of the loan issued |
| annual_inc | Numeric (continuous) | Annual income of the borrower |
| emp_length | Categorical | Employment length in years (0–10) |
| dti | Numeric (continuous) | Debt-to-income ratio |
| grade | Ordinal factor (A–G) | Loan grade assigned by Lending Club; A is best, G is worst |
| purpose | Nominal factor | Stated purpose of the loan |
| home_ownership | Nominal factor | Borrower's home ownership status |
| addr_state | Nominal factor | State of the borrower |
| term | Ordinal factor | Length of the loan (36 or 60 months) |


### Data Quality Report

```{r}
glimpse(actual_data)
dim(actual_data)
DataExplorer::introduce(actual_data)
DataExplorer::plot_missing(actual_data)
missing_tbl <- actual_data |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Feature", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

```{r}
any(duplicated(actual_data))
```

##Numerical Features
```{r}
numeric_features <- actual_data |>
  select_if(is.numeric)

num_s <- skimr::skim_without_charts(numeric_features)

focus_numeric_table <- as.data.frame(num_s) |>
  mutate(count = nrow(actual_data) - n_missing) |>
  select(skim_variable, count, numeric.p0, numeric.p25, numeric.mean, 
         numeric.p50, numeric.p75, numeric.p100, numeric.sd)

extra_two <- funModeling::status(numeric_features)[ ,c("p_na", "unique")]
rownames(extra_two) <- NULL

numeric_report_table <- cbind(focus_numeric_table, extra_two) |>
  relocate(p_na, .before = 3) |>
  relocate(unique, .before = 4)

names(numeric_report_table) <- c("Feature", "Count", "% Miss.", "Cardinality", "Min", "1st Qrt.", "Mean", "Median", "3rd Qrt.", "Max", "Std. Dev.")

numeric_report_table
```
```{r}
funModeling::plot_num(numeric_features)
funModeling::profiling_num(numeric_features)
```


```{r}
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  sum(x < lower | x > upper, na.rm = TRUE)
}

actual_data |>
  select(annual_inc, int_rate, loan_amnt, dti) |>
  summarise(across(everything(), detect_outliers)) |>
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_outliers") |>
  gt()
```


#### Categorical Features

```{r}
categorical_features <- actual_data |>
  select_if(is.factor)

cat_s <- skim_without_charts(categorical_features)

focus_cat_table <- as.data.frame(cat_s) |>
  mutate(count = nrow(actual_data) - n_missing) |>
  mutate(percent_Miss = round((1 - complete_rate) * 100, 1)) |>
  select(skim_variable, count, percent_Miss, factor.n_unique, factor.top_counts)

names(focus_cat_table) <- c("Feature", "Count", "% Miss.", "Card.", "Top Counts")

focus_cat_table
funModeling::freq(categorical_features)
```

### Data Issues and Proposed Handling Techniques

The following issues were identified in the raw dataset and addressed during data cleaning:

| Feature | Data Quality Issue | Proposed Handling Solution |
| :--- | :--- | :--- |
| annual_inc | 4 missing values (0.0005%) | Retain; filter within chunks requiring it |
| annual_inc | 39,719 outliers; max $9,500,000 plausible | Valid outliers; apply clamp transformation |
| int_rate | 6,308 outliers; range 5.32–28.99% realistic | Valid outliers; apply clamp transformation |
| loan_amnt | No missing values; 0 outliers | No action needed |
| dti | 81 outliers; max 9,999 likely invalid | Remove rows where dti > 100 |
| emp_length | n/a present as level (44,825 / 5.05%) | Recoded to NA via str_squish and na_if |
| addr_state | Cardinality 51 — expected (50 states + DC) | No action needed |
| grade | Skewed toward B and C (56% combined) | Valid distribution; no action needed |
| purpose | Rare categories <0.1% each | Consider collapsing into other |
| home_ownership | OTHER/NONE/ANY represent <0.03% | Collapse into single OTHER category |
| term | 70/30 split — imbalanced but realistic | Valid distribution; no action needed |
| All character columns | Inconsistent spacing and blank-like values | str_squish, na_if, case_when standardization |



### Data Cleaning
To see the basic format of the data and get a sense of how many rows and columns we have.
```{r}
glimpse(actual_data)
dim(actual_data)
```

```{r}
(missing_tbl <- actual_data |>
    summarise(across(everything(), ~ sum(is.na(.)))) |>
    pivot_longer(everything(), names_to = "Feature", values_to = "n_missing") |>
    arrange(desc(n_missing)))
```

Check to see if there are any duplicated data rows:
```{r}
any(duplicated(actual_data))
```


Removing all the duplicates and keeping the first occurrence:
```{r}
actual_data <- actual_data[!duplicated(actual_data), ]
```

Cleaning string variables
```{r}
actual_data <- actual_data |>
  mutate(across(where(is.character), ~ str_squish(.))) |>
  mutate(across(where(is.character), ~ na_if(.x, ""))) |>
  mutate(across(where(is.character),
                ~ case_when(
                  .x %in% c("NA", "N/A", "null", "NULL", "?", "unknown", "Unknown") ~ NA_character_,
                  TRUE ~ .x
                )))
```


```{r}
actual_data <- actual_data |>
  mutate(
    grade      = factor(grade, levels = c("A","B","C","D","E","F","G"), ordered = TRUE),
    term       = factor(term,  levels = c("36 months","60 months"), ordered = TRUE),
    emp_length = factor(emp_length,
                   levels = c("< 1 year", "1 year", "2 years", "3 years",
                              "4 years", "5 years", "6 years", "7 years",
                              "8 years", "9 years", "10+ years"),
                   ordered = TRUE),
    purpose    = fct_relabel(purpose, ~ str_replace_all(.x, "_", " "))
  )
```


Drop rows missing key variables:
```{r}
actual_data <- actual_data |>
  drop_na()
```

```{r}
(missing_tbl_final <- actual_data |>
    summarise(across(everything(), ~ sum(is.na(.)))) |>
    pivot_longer(everything(), names_to = "Feature", values_to = "n_missing") |>
    arrange(desc(n_missing)))
```



# Creator: Tracy Huang

### Q1: Can interest rate by predicted based on loan term and loan grade?

Summary:
```{r}
q1_table <- actual_data |>
  group_by(grade, term) |>
  summarise(
    n = n(),
    mean_int_rate = round(mean(int_rate, na.rm = TRUE), 2),
    median_int_rate = round(median(int_rate, na.rm = TRUE), 2),
    sd_int_rate = round(sd(int_rate, na.rm = TRUE), 2),
    q1 = round(quantile(int_rate, 0.25, na.rm = TRUE), 2),
    q3 = round(quantile(int_rate, 0.75, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(grade, term)

q1_table |> gt()
```

Graph:
```{r}
actual_data |>
  ggplot(aes(x = grade, y = int_rate, fill = term)) +
  geom_boxplot() +
  labs(
    title = "Interest Rate by Loan Grade and Term",
    x = "Loan Grade",
    y = "Interest Rate (%)",
    fill = "Term"
  ) +
  theme_minimal()
```

Interpretation of Findings:



### Q2: Can home ownership status be predicted using borrower income and employment length?

Summary:
```{r}
q2_table <- actual_data |>
  mutate(inc_decile = ntile(annual_inc, 10)) |>
  group_by(home_ownership, inc_decile) |>
  summarise(
    n = n(),
    inc_min = round(min(annual_inc, na.rm = TRUE), 0),
    inc_max = round(max(annual_inc, na.rm = TRUE), 0),
    mean_inc = round(mean(annual_inc, na.rm = TRUE), 0),
    .groups = "drop"
  ) |>
  arrange(home_ownership, inc_decile)

q2_table |> gt()
```


Graph: 
```{r}
q2_plot_data <- actual_data |>
  mutate(inc_decile = ntile(annual_inc, 10)) |>
  group_by(home_ownership, inc_decile) |>
  summarise(
    median_inc = median(annual_inc, na.rm = TRUE),
    q1 = quantile(annual_inc, 0.25, na.rm = TRUE),
    q3 = quantile(annual_inc, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(q2_plot_data, 
       aes(x = inc_decile, y = median_inc, 
           group = home_ownership, color = home_ownership)) +
  geom_line() +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q1, ymax = q3), width = 0.2) +
  labs(
    title = "Median Annual Income Across Deciles by Home Ownership",
    x = "Income Decile (1 = Lowest, 10 = Highest)",
    y = "Median Annual Income",
    color = "Home Ownership"
  ) +
  theme_minimal()
```
Interpretation of Findings:


# Interpreter: Kyran Lakhia

### Q1: Which states are experiencing the lowest grades on average?

```{r}
# Summary table: average grade by state (bottom 10)
state_avg <- actual_data |>
  mutate(grade_num = as.numeric(grade)) |>
  group_by(addr_state) |>
  summarise(avg_grade = mean(grade_num, na.rm = TRUE)) |>
  arrange(desc(avg_grade)) |>
  slice_head(n = 10) |>
  mutate(avg_grade_label = c("A","B","C","D","E","F","G")[round(avg_grade)])

state_avg
```

```{r}
# Bar chart: states with worst average loan grades
ggplot(state_avg, aes(x = reorder(addr_state, avg_grade), y = avg_grade, fill = avg_grade)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_gradient(low = "#fcae91", high = "#cb181d") +
  scale_y_continuous(
    breaks = 1:7,
    labels = c("A","B","C","D","E","F","G")
  ) +
  labs(
    title = "States with the Worst Average Loan Grades",
    subtitle = "Loan grades range from A (best) to G (worst)",
    x = "State",
    y = "Average Loan Grade"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold")
  )
```

The chart shows the states with the worst average loan grades. Since loan grades range from A (best) to G (worst), higher numeric averages indicate lower credit quality. However, the averages are still between B and C, which suggests that overall loan quality remains fairly strong even in the states with the lowest average grades.

### Q2: Is there a relationship between home ownership and purpose?

```{r}
# Summary table: contingency table of home ownership x purpose
table(actual_data$home_ownership, actual_data$purpose)
```

```{r}
# Stacked bar chart: loan purpose by home ownership status
actual_data |>
  filter(home_ownership %in% c("RENT", "MORTGAGE", "OWN")) |>
  group_by(home_ownership, purpose) |>
  summarise(count = n(), .groups = "drop") |>
  ggplot(aes(x = home_ownership, y = count, fill = purpose)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Loan Purpose by Home Ownership Status",
    x = "Home Ownership Status",
    y = "Number of Loans",
    fill = "Loan Purpose"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Orator 1: Ross Guinan

### Q1: 

Summary:
```{r}
#
```


Graph:
```{r}

```

Interpretation of Findings:
```{r}

```

### Q2: 

Summary:
```{r}
#
```


Graph:
```{r}

```

Interpretation of Findings:
```{r}

```

# Orator 2: Carter Boxell

```{r}
lc_data <- read_csv("lc_2016_2017.csv")
lc_data$grade <- factor(lc_data$grade, levels = c("A", "B", "C", "D", "E", "F", "G"))
lc_data$addr_state <- factor(lc_data$addr_state)
lc_data$purpose <- factor(lc_data$purpose)
```

### Q1: How do the average interest rates for each grade compare?

Summary:
```{r}
#
```


Graph:
```{r}
ggplot(
  dplyr::summarize(
    dplyr::group_by(lc_data, grade),
    avg_int_rate = mean(int_rate, na.rm = TRUE)
  ),
  aes(x = grade, y = avg_int_rate)
) +
  geom_col() +
  labs(
    title = "Average Interest Rate by Loan Grade",
    x = "Loan Grade",
    y = "Average Interest Rate"
  )
```

Interpretation of Findings:
```{r}

```

### Q2: Is there a relationship between state and purpose of loan?

Summary:
```{r}
#
```


Graph:
```{r}
ggplot(lc_data, aes(x = addr_state, fill = purpose)) +
  geom_bar(position = "fill") +
  labs(
    title = "Proportion of Loan Purposes by State",
    x = "State",
    y = "Proportion of Loans",
    fill = "Loan Purpose"
  )
```

Interpretation of Findings:
```{r}

```


# Deliverer: Emily Jones

### Q1: 

Summary:
```{r}
#
```


Graph:
```{r}

```

Interpretation of Findings:
```{r}

```

### Q2: 

Summary:
```{r}
#
```


Graph:
```{r}

```

Interpretation of Findings:
```{r}

```



# Follow-up Questions

### New Questions Based Off Initial Investigation

- Q1: WRITE_QUESTION_HERE
- Q2: WRITE_QUESTION_HERE
- Q3: WRITE_QUESTION_HERE 
- Q4: WRITE_QUESTION_HERE

### Investigation of Follow-up Questions

GIVE WHAT 2 QUESTIONS YOU ATTEMPTED TO INVESTIGATE FURTHER IN COMPLETE SENTENCES (*Example:* Our group decided to investigate Q2 and Q4 in further detail.)

SHOW AT LEAST 2 TABLES OR FIGURES BELOW THAT EXPLORE ANSWERS FOR THE QUESTIONS YOU ARE INVESTIGATING FURTHER.

```{r}
#
```

```{r}
#
```

# Summary

GIVE A 2 PARAGRAPH SUMMARY. 

PARAGRAPH 1 SHOULD DESCRIBE WHAT YOU LEARNED ABOUT YOUR DATA FROM INVESTIGATING THE INITIAL QUESTIONS. DID YOU FIND ANYTHING UNUSUAL IN YOUR DATA? DID ANYTHING SURPRISE YOU? WHICH OF THE INITIAL QUESTIONS WERE HELPFUL IN LEADING YOU TO MORE QUESTIONS?

PARAGRAPH 2 SHOULD SUMMARIZE WHAT YOU LEARNED FROM INVESTIGATING THE FOLLOW-UP QUESTIONS. WHY ARE THESE FOLLOW-UP QUESTIONS INTERESTING FOR INVESTIGATION? DESCRIBE THE TABLES/FIGURES YOU USED TO EXPLORE ANSWERS TO THESE FOLLOW-UP QUESTIONS? WHAT DID YOU LEARN FROM THE TABLES/FIGURES REGARDING THE FOLLOW-UP QUESTIONS YOU PROPOSED?